Detection of Oscillations in Aldebaran with Ground-Based Observations

### 2018-01-19 paper/aldebaran.tex
Several updates around results and archival data section.
* Some wordsmithing and clarity in time-series and CARMA appendix.

* Plot of nu_max posteriors from archival, SONG, and combined data sets.

* More numbers (nu_max and Q posterior ranges) in archival, SONG, and
combined sections.

@@

Fortunately, this GP can be written as the solution of a class of stochastic ordinary differential equations, the Continuous Auto-Regressive Moving Average (CARMA) models, whose likelihoods can be evaluated in linear time. For problems such as the damped driven harmonic oscillator, we therefore have access to these powerful computational tools for inferring their power spectral densities. \citet{Kelly2014} gives more details on CARMA
processes in an astronomical context, and shows how a state-space model of the solution
$y(t)$ can be tracked through a time-series of uncertain observations $y_k =
y\left( t_k \right) + \epsilon_k$, with Gaussian noise $\epsilon_k$ of zero mean
and known (heteroskedastic) variance, using a Kalman filter to produce a
likelihood function $p\left( y_k \mid \sigma, r_i, b_j \right)$ in
$\mathcal{O}\left( n \right)$ computational time.  Additionally,
\citet{Foreman-Mackey2017} show how a dimensional expansion can be used to
reduce the standard Gaussian Process covariance matrix for the process with
autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under observations at times
$t_k$ to a banded form that can then be Cholesky-decomposed in the standard
Gaussian Process likelihood function in $\mathcal{O}\left( n \right)$ time.

Here we have adopted the \citet{Kelly2014} approach.  The \texttt{CARMA.jl} package\footnote{\url{https://github.com/farr/CARMA.jl}} implements Kalman filters that can compute the likelihood for a set of observations parameterised by either the roots $r_i$ and $b_j$ or the RMS amplitudes $A_i$ and roots $r_i$.  We also include a Keplerian RV signal from a planetary companion to Aldebaran \citep{Hatzes2015,Gregory2005}.  Because we have a set of observations taken with different instruments at different sites (see Section \ref{sec:archival-obs}), we include RV offset (i.e.\ mean value) parameters that are instrument- and site-dependent and also an instrument- and site-dependent scaling parameter.  Thus, the $i$th RV measured at site and intstrument combination $j$, $y_{ij}$, is assumed to be
\begin{equation}
  \label{eq:RV-model}
  y_{ij} = v_r\left(t_i\right) + \mu_j + y\left( t_i \right) + \nu_j \epsilon_{ij},
\end{equation}
$t_i$ is the time of the measurement, $v_r$ is the planetary RV signal, $\mu_j$ is the mean RV offset for site and instrument combination $j$, $y(t)$ is the CARMA process representing stellar activity, $\nu_j$ is the uncertainty scaling parameter for site and instrument combination $j$, and $\epsilon_{ij}$ is an independent Gaussian random variable with mean zero and standard deviation equal to the quoted uncertainty of the observation.  We impose broad priors (see Table \ref{tab:parameters}) on the parameters of the planetary RV signal, the CARMA process, the $\mu_j$ offsets, and the $\nu_j$ scale factors.  We treat the frequency of maximum asteroseismic power,  $\nu_\mathrm{max}$, as the frequency of an appropriate oscillatory mode in the CARMA process representing the stellar activity; this is equivalent to fitting a Lorentzian profile to a set of unresolved asteroseismic modes.  We used the nested sampling algorithm of the \texttt{Ensemble.jl} package\footnote{\url{https://github.com/farr/Ensemble.jl}; this package implements several stochastic sampling algorithms based on the ``stretch move'' proposal used in \texttt{emcee} \citep{Foreman-Mackey2013} and introduced in \citet{Goodman2010}} to calculate the marginal likelihood (evidences) and draw samples from the posterior distribution over the parameters of our various models.

@@

\citet{Hatzes2015} reported on RV observations of Aldebaran from the coud\'{e}
\'{e}chelle spectrograph of the Th\"{u}ringer Landessternwarte Tautenburg (TLS), the
Tull Spectrograph of the McDonald 2.7 m telescope, and the Bohyunsan Observatory
\'{E}chelle Spectrograph (BOES) spectrograph of the Bohyunsan Optical Astronomy
Observatory (BOAO) spanning the period from 2000.01 to 2013.92.  Combined with earlier observations \citep{hatzes1993}, we construct a full time series spanning more than two decades (Figure~\ref{alldata}) and attempt to fit this with a Keplerian model and a CARMA noise process.

@@

\subsection{SONG Observations}    % this goes into section 4.1

In order to confirm the oscillations detected in the archival data, we used the Hertzsprung SONG telescope \citep{2017ApJ...836..142G} to conduct high time cadence follow-up observations. These were carried out in the highest resolution mode ($R\approx110000$)
with an integration time of 30\,s, between 2016 September 27 and 2017 January 12. During this time we attempted to obtain at least one
visit per available night, for a total of 254 epochs over the campaign. The spectra were
extracted using the SONG pipeline \citep[see][]{2017ApJ...836..142G}. SONG employs an iodine cell for
precise wavelength calibration and the velocities were determined using the \textit{iSONG}
software following the same procedures as described in \cite{2017ApJ...836..142G}. The typical
velocity precision per visit is 2.5$\,\rm m\,s^{-1}$ allowing us to easily detect the oscillations
which displays a peak-to-peak amplitude of $\approx \, 170 \, \rm m \, s^{-1}$. Figure~\ref{songdata} displays
the radial velocities obtained -- oscillations are easily visible as well as a long-term
trend, probably due to the planetary companion.

@@

By jointly fitting a Keplerian for the planet with a CARMA model for the stellar oscillations, we obtain a precise estimate of both \numax and the planetary orbital parameters. The planetary orbital parameters we obtain are similar to \citet{Hatzes2015}: period $P = 629.0^{+ 1.7}_{-2.1}$~d, eccentricity $e = 0.168^{+ 0.069}_{- 0.071}$; and radial velocity semi-amplitude $K = 127^{+ 15}_{- 14}$~m/s.

The stellar oscillations are best fit with a single CARMA component with $\numax = 2.238^{+ 0.086}_{- 0.080}$ \muHz, and including additional CARMA components does not resolve any further structure in the power spectrum. We have determined the stellar properties of Aldebaran with the additional constraint of \numax determined from this analysis.  The details of the stellar modelling are presented in detail in Appendix \ref{sec:mod}.  We have considered the impact of our additional \numax constraint and run our analysis for different sets of spectroscopic estimates.  For final values we adopt the \citet{2012Sheffield} spectroscopic solution as being `middle of the road' estimates.  For the analysis without constraint on \numax we find estimates of the stellar properties as $M = 1.27^{+0.24}_{-0.2} \, \mathrm{M_{\odot}}$ and age $4.86^{+3.56}_{-2.04} \, \rm Gyr$.  With the inclusion of \numax we find $M = 1.16^{+0.07}_{-0.07} \, \mathrm{M_{\odot}}$ and age $6.38^{+1.42}_{-1.12} \, \rm Gyr$.  It is clear that the addition of \numax provides substantially more precise estimates of mass and age. With this stellar mass, the radial velocity signal translates to an inclination-dependent planetary mass of $M\,\sin\,{i} = 5.81 \pm 0.72~M_J$ \citep[Equation~1]{Torres2008}, a massive giant planet; for sufficiently high inclinations ($\gtrsim 45^\circ$), the mass of Aldebaran~b may exceed $13~M_J$, making it a possible low-mass brown dwarf candidate.  % re sheffield - I have substituted the citation from the .bib but you said 'b'; afaik she only published one paper in 2012? - Ben

@@

@@

A Gaussian Process such as that used by \citet{brewer2009} also corresponds to
the solution, $y(t)$, of the stochastic ODE \begin{equation}
\label{eq:GP-definition} \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) =
\prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t), \end{equation} where
$p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the process; $r_i \in
\mathbb{C}$, $\Re r_i < 0$ are the roots of the characteristic equation of the
ODE; $b_j \in \mathbb{C}$, $\Re b_j < 0$ are corresponding roots of the inverse
process; and $\eta(t)$ is a white-noise Gaussian Process with $\left\langle
\eta(t) \right\rangle = 0$ and $\left\langle \eta(t) \eta(t') \right\rangle =
\sigma^2 \delta\left(t - t' \right)$.  Such a solution is called a CARMA
(Continuous AutoRegressive Moving Average) process.  The constraint that $\Re
r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator defining the process
and its inverse are invertable.  Under the assumption that the solution, $y(t)$,
is real, roots $r_i$ and $b_j$ must either be real or occur in complex conjugate
pairs.  The power spectrum of the process is \begin{equation}
\label{eq:CARMA-PSD} P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i
f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation} The power spectrum is a rational function of frequency with poles
at $2\pi i f = r_i$ and zeros at $2\pi i f = b_i$; because of the invertable
constraints, these poles and zeros all occur when $\Im f \neq 0$.  The
autocorrelation function corresponding to the power spectrum in Eq.\
\eqref{eq:CARMA-PSD} is \begin{equation} \label{eq:CARMA-ACF} \rho_y (\tau) =
\left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle =
\sum_{i=1}^p A_i^2 e^{r_i \tau}, \end{equation} where the RMS amplitudes of the
$p$ different modes, $A_i$, are functions of $\sigma$ and the values of the
roots $r_i$ and $b_j$.  If $q = p-1$, then the $A_i$ are independent, while if
$q < p-1$ then there are interdependencies among the $A_i$.  From either the
power spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the autocorrelation function in
Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a real root $r_i$ represents an
exponentially-decaying mode with $r_i$ the rate constant, while a complex
conjugate pair of $r_i$ represent an oscillatory mode with decay rate $\Re r_i$
and angular frequency $\left|\Im r_i\right|$.  Oscillatory modes can also be
described by their frequency $f$ and quality factor $Q$ via $r = f/(2Q) \pm 2\pi
i f$.  Decaying modes generate a damped random walk, also known as a
Ornstein-Uhlenbeck process.

>>>>

% Numbers in the document
\newcommand{\twosidedrange}[3]{\ensuremath{#1^{+#2}_{-#3}}}

%% Q
\newcommand{\HatzesQRange}{\twosidedrange{2.12}{0.69}{0.56}}
\newcommand{\SONGQRange}{\twosidedrange{3.06}{1.20}{0.81}}
\newcommand{\CombinedQRange}{\twosidedrange{2.83}{0.60}{0.51}}

%% NuMax
\newcommand{\HatzesNuMaxRange}{\twosidedrange{2.21}{0.13}{0.12}}
\newcommand{\SONGNuMaxRange}{\twosidedrange{2.33}{0.13}{0.14}}
\newcommand{\combinedNuMaxRange}{\twosidedrange{2.238}{0.086}{0.080}}

%% P
\newcommand{\PeriodRange}{\twosidedrange{629.0}{1.7}{2.1}}

%% e
\newcommand{\EccentricityRange}{\twosidedrange{0.168}{0.069}{0.071}}

%% K
\newcommand{\PlanetRVAmpRange}{\twosidedrange{127}{15}{14}}

@@

Fortunately, this GP can be written as the solution of a class of stochastic
ordinary differential equations, the Continuous Auto-Regressive Moving Average
(CARMA) models, whose likelihoods can be evaluated in linear time. For problems
such as the damped driven harmonic oscillator, we therefore have access to these
powerful computational tools for inferring their power spectral densities.
\citet{Kelly2014} gives more details on CARMA processes in an astronomical
context, and shows how a state-space model of the solution $y(t)$ can be tracked
through a time-series of uncertain observations $y_k = y\left( t_k \right) +
\epsilon_k$, with Gaussian noise $\epsilon_k$ of zero mean and known
(heteroskedastic) variance, using a Kalman filter to produce a likelihood
function $p\left( y_k \mid \sigma, r_i, b_j \right)$ depending on the amplitude
of the stochastic forcing, $\sigma$, the (possibly complex) eigenfrequencies of
the ODE, $r_i$, and parameters describing the power spectrum of the stochastic
forcing, $b_j$ in $\mathcal{O}\left( n \right)$ computational time.
Additionally, \citet{Foreman-Mackey2017} show how a dimensional expansion can be
used to reduce the standard Gaussian Process covariance matrix for this same
process,  with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$, to a banded form that can then be
Cholesky-decomposed in the standard Gaussian Process likelihood function in
$\mathcal{O}\left( n \right)$ time.

Here we have adopted the \citet{Kelly2014} approach.  The \texttt{CARMA.jl}
package\footnote{\url{https://github.com/farr/CARMA.jl}} implements Kalman
filters that can compute the likelihood for a set of observations parameterised
by either the $r_i$ and $b_j$ or the RMS amplitudes $A_i$ of each eigenmode of
the ODE and the corresponding roots $r_i$.  We also include a Keplerian RV
signal from a planetary companion to Aldebaran \citep{Hatzes2015,Gregory2005}.
Because we have a set of observations taken with different instruments at
different sites (see Section \ref{sec:archival-obs}), we include RV offset
(i.e.\ mean value) parameters that are instrument- and site-dependent and also
an instrument- and site-dependent scaling parameter.  Thus, the $i$th RV
measured at site and intstrument combination $j$, $y_{ij}$, is assumed to be
%
\begin{equation}
  \label{eq:RV-model}
  y_{ij} = v_r\left(t_i\right) + \mu_j + y\left( t_i \right) + \nu_j
\epsilon_{ij},
\end{equation}
%
where $t_i$ is the time of the measurement, $v_r$ is the planetary RV signal,
$\mu_j$ is the mean RV offset for site and instrument combination $j$, $y(t)$ is
the CARMA process representing stellar activity, $\nu_j$ is the uncertainty
scaling parameter for site and instrument combination $j$, and $\epsilon_{ij}$
is an independent Gaussian random variable with mean zero and standard deviation
equal to the quoted uncertainty of the observation.  We impose broad priors (see
Table \ref{tab:parameters}) on the parameters of the planetary RV signal, the
CARMA process, the $\mu_j$ offsets, and the $\nu_j$ scale factors.  We treat the
frequency of maximum asteroseismic power, $\nu_\mathrm{max}$, as the frequency
of an appropriate eigenmode in the CARMA process representing the stellar
activity; this is equivalent to fitting a Lorentzian profile to a set of
unresolved asteroseismic modes.  We used the nested sampling algorithm of the
\texttt{Ensemble.jl} package\footnote{\url{https://github.com/farr/Ensemble.jl};
this package implements several stochastic sampling algorithms based on the
``stretch move'' proposal used in \texttt{emcee} \citep{Foreman-Mackey2013} and
introduced in \citet{Goodman2010}} to calculate the marginal likelihood
(evidences) and draw samples from the posterior distribution over the parameters
of our various models.  The priors we impose on the parameters of our model are given in Table \ref{tab:parameters}.

@@

\citet{Hatzes2015} reported on RV observations of Aldebaran from the coud\'{e}
\'{e}chelle spectrograph of the Th\"{u}ringer Landessternwarte Tautenburg (TLS),
the Tull Spectrograph of the McDonald 2.7 m telescope, and the Bohyunsan
Observatory \'{E}chelle Spectrograph (BOES) spectrograph of the Bohyunsan
Optical Astronomy Observatory (BOAO) spanning the period from 2000.01 to
2013.92.  Combined with earlier observations \citep{hatzes1993}, we construct a
full time series spanning more than two decades from 1980.80 to 2013.71
(Figure~\ref{alldata}) and fit this with a Keplerian model and a CARMA process.
We find evidence for a low-quality ($Q = \HatzesQRange{}$) oscillatory mode with
$\nu_\mathrm{max} = \HatzesNuMaxRange{} \, \mu\mathrm{Hz}$.  This is consistent
with the presence of a number of un-resolved asteroseismic modes in the RV data.
Our posterior for \numax{} from this data set is shown in Figure
\ref{fig:numax-datasets}.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../plots/numax}
  \caption{Posterior for \numax{} from the complete, combined RV data set described in this paper (blue), the archival data from \citet{Hatzes2015} (green), and the SONG data described in Section \ref{sec:SONG-observations} (red).  The inference on the archival data is consistent with the independent inference on the SONG data set alone.}
  \label{fig:numax-datasets}
\end{figure}

\subsection{SONG Observations}    % this goes into section 4.1
\label{sec:SONG-observations} In order to confirm the oscillations detected in
the archival data, we used the Hertzsprung SONG telescope
\citep{2017ApJ...836..142G} to conduct high time cadence follow-up observations.
These were carried out in the highest resolution mode ($R\approx110000$) with an
integration time of 30\,s, between 2016 September 27 and 2017 January 12. During
this time we attempted to obtain at least one visit per available night, for a
total of 254 epochs over the campaign. The spectra were extracted using the SONG
pipeline \citep[see][]{2017ApJ...836..142G}. SONG employs an iodine cell for
precise wavelength calibration and the velocities were determined using the
\textit{iSONG} software following the same procedures as described in
\cite{2017ApJ...836..142G}. The typical velocity precision per visit is
2.5$\,\rm m\,s^{-1}$ allowing us to easily detect the oscillations which
displays a peak-to-peak amplitude of $\approx \, 170 \, \rm m \, s^{-1}$.
Figure~\ref{songdata} displays the radial velocities obtained -- oscillations
are easily visible as well as a long-term trend, probably due to the planetary
companion.  From this data set alone we also find evidence for a low-quality ($Q
= \SONGQRange{}$) oscillatory mode with $\nu_\mathrm{max} = \SONGNuMaxRange{} \,
\mu\mathrm{Hz}$ which is consistent with the inference from the archival data.
This posterior is shown in Figure \ref{fig:numax-datasets}.

@@

By jointly fitting a Keplerian for the planet with a CARMA model for the stellar
oscillations to the combined archival and SONG data sets (Figure \ref{alldata}),
we obtain a precise estimate of both \numax and the planetary orbital
parameters. The planetary orbital parameters we obtain are similar to
\citet{Hatzes2015}, but with larger uncertainty that is likely due to our
more-flexible and correlated model for the stellar component of the RV signal:
period $P = \PeriodRange \, \mathrm{d}$, eccentricity $e =
\EccentricityRange{}$; and radial velocity semi-amplitude $K =
\PlanetRVAmpRange{} \, \mathrm{m} \, \mathrm{s}^{-1}$.

The stellar oscillations are best fit with a single, low-quality ($Q =
\CombinedQRange{}$) oscillatory mode with $\numax = 2.238^{+ 0.086}_{- 0.080}$
\muHz.  We find no improvement in the marginal likelihood (evidence) or other
model-selection information criteria \citep{Gelman2013} from models with
additional oscillatory modes, so we conclude that the large spacing $\Delta\nu$
is not observable in this data set.  We have determined the stellar properties
of Aldebaran with the additional constraint of \numax determined from this
analysis.  The details of the stellar modelling are presented in Appendix
\ref{sec:mod}.  We have considered the impact of our additional \numax
constraint and run our analysis for different sets of spectroscopic estimates.
For final values we adopt the \citet{2012Sheffield} spectroscopic solution as
being `middle of the road' estimates.  For the analysis without constraint on
\numax we find estimates of the stellar properties as $M = 1.27^{+0.24}_{-0.2}
\, \mathrm{M_{\odot}}$ and age $4.86^{+3.56}_{-2.04} \, \rm Gyr$.  With the
inclusion of \numax we find $M = 1.16^{+0.07}_{-0.07} \, \mathrm{M_{\odot}}$ and
age $6.38^{+1.42}_{-1.12} \, \rm Gyr$.  It is clear that the addition of \numax
provides substantially more precise estimates of mass and age. With this stellar
mass, the radial velocity signal translates to an inclination-dependent
planetary mass of $M\,\sin\,{i} = 5.81 \pm 0.72~M_J$
\citep[Equation~1]{Torres2008}, a massive giant planet; for sufficiently high
inclinations ($\gtrsim 45^\circ$), the mass of Aldebaran~b may exceed $13~M_J$,
making it a possible low-mass brown dwarf candidate.

@@

The data, code, and \LaTeX{} source used to produce this document can be found
at \url{https://github.com/farr/Aldebaran} under an open-source ``MIT'' license.

@@

A CARMA process corresponds to the solution, $y(t)$, of the stochastic ODE
%
\begin{equation}
  \label{eq:GP-definition}
  \prod_{i=1}^{p} \left[ \diff{}{t} - r_i \right] y(t) =
  \prod_{j = 1}^{q} \left[ \diff{}{t} - b_i \right] \eta(t),
\end{equation}
%
where $p, q \in \mathbb{Z}$, $p > q \geq 0$ define the order of the process;
$r_i \in \mathbb{C}$, $\Re r_i < 0$ are the roots of the characteristic equation
of the ODE; $b_j \in \mathbb{C}$, $\Re b_j < 0$ are corresponding roots of the
inverse process; and $\eta(t)$ is a white-noise Gaussian Process with
$\left\langle \eta(t) \right\rangle = 0$ and $\left\langle \eta(t) \eta(t')
\right\rangle = \sigma^2 \delta\left(t - t' \right)$.  The constraint that $\Re
r_i < 0$ and $\Re b_j < 0$ ensures that the linear operator defining the process
and its inverse are invertable.  Under the assumption that the solution, $y(t)$,
is real, roots $r_i$ and $b_j$ must either be real or occur in complex conjugate
pairs.  The power spectrum of the process is
%
\begin{equation}
  \label{eq:CARMA-PSD}
  P_y(f) = \sigma^2 \left| \frac{\prod_{j=1}^q \left[ 2 \pi i f - b_j \right]}{\prod_{i=1}^p \left[ 2 \pi i f - r_i \right]} \right|^2.
\end{equation}
%
The power spectrum is a rational function of frequency with poles at $2\pi i f =
r_i$ and zeros at $2\pi i f = b_i$; because of the invertable constraints, these
poles and zeros all occur when $\Im f \neq 0$.  The autocorrelation function
corresponding to the power spectrum in Eq.\ \eqref{eq:CARMA-PSD} is
%
\begin{equation}
  \label{eq:CARMA-ACF}
  \rho_y (\tau) = \left\langle y\left( t \right) y \left( t + \tau \right) \right\rangle = \sum_{i=1}^p A_i^2 e^{r_i \tau},
\end{equation}
%
where the RMS amplitudes of the $p$ different modes, $A_i$, are functions of
$\sigma$ and the values of the roots $r_i$ and $b_j$.  If $q = p-1$, then the
$A_i$ are independent, while if $q < p-1$ then there are interdependencies among
the $A_i$.  From either the power spectrum in Eq.\ \eqref{eq:CARMA-PSD} or the
autocorrelation function in Eq.\ \eqref{eq:CARMA-ACF}, it is apparent that a
real root $r_i$ represents an exponentially-decaying mode with $r_i$ the rate
constant, while a complex conjugate pair of $r_i$ represent an oscillatory mode
with decay rate $\Re r_i$ and angular frequency $\left|\Im r_i\right|$.
Oscillatory modes can also be described by their frequency $f$ and quality
factor $Q$ via $r = f/(2Q) \pm 2\pi i f$.  Decaying modes generate a damped
random walk, also known as a Ornstein-Uhlenbeck process.  More discussion of CARMA processes, including formulae for $A_i$ in terms of $\sigma$, $r_i$, and $b_j$, can be found in \citet{Kelly2014} and references therein.

The priors we impose on the parameters of the CARMA model, planetary RV signal,
and instrumental and site effects are given in Table \ref{tab:parameters}.
